Data exploration, quality control and statistical analysis of ChIP-exo experiments

### 2016-05-03 rmarkdown/paper.Rnw
fixed the qc tables and finished background

\usepackage{xcolor}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{float}
\usepackage{verbatim}

@@

@@

\begin{enumerate}
\item Available data, the motivation of this part is two fold:
  \begin{itemize}
  \item we want to list all data sets we obtained
  \item we want to emphasize that for the sig70 datasets we have a
    unique design that allows us to compare chip exo with SE and PE
    chipseq , the later one for the first time
  \end{itemize}
\item comparison with chip seq ? with and without dpeak ?
\item only in methods. list the align algorithm used
\item chip seq qc metrics
\item chip exo qc pipeline, description and justification of the steps
\begin{itemize}
\item need to justify the sampling step. for this we may use the depth
  table and add the factor to normalize one depth to another. explain
  the confounding of pcr amplification and the reduction of the number
  of possible mappable positions.
\item case study with FoxA1, analysis of enrichment vs library
  complexity and analysis and the strand imbalance tools. in methods
  we need to add which tests did we perform to show that the pipeline
  works. how do we expect from the diagnostics and how to interpret
  the plots.
\item analysis of why the pipeline works. in methods need to specify
  how and in results need to specify why it works, and what are the
  conclusions from it
\end{itemize}
\item high resolution analysis with chip exo and comparison with PE
  chip seq
  \begin{itemize}
  \item justify what do we used dpeak (in methods need to list the
    other methods used). perhaps we can add gof plots in
    supplement. we also want to compare with the same method,
    therefore we can use the fact that dpeak have a unified model for
    PE and SE dpeak
  \item after comparing with other methods we can compare with PE and
    SE ChIP-seq, this comparison gives also resolution and sensitivity
  \item compare samples under same budget of reads
  \end{itemize} 
\end{enumerate}

@@

\section{Background}
\label{sec:intro}

ChIP-exo (Chromatin Immunoprecipitation followed by exonuclease
digestion and next generation sequencing) Rhee and Pugh, 2011
\cite{exo1} is the state-of-the-art experiment developed to attain
single base-pair resolution of protein binding site identification and
it is considered as a powerful alternative to popularly used ChIP-Seq
(Chromatin Immunoprecipitation coupled with next generation
sequencing) assay. ChIP-exo experiments first capture millions of DNA
fragments (150 - 250 bp in length) that the protein under study
interacts with using random fragmentation of DNA and a
protein-specific antibody. Then, exonuclease is introduced to trim the
$5\prime$ end of each DNA fragment to a fixed distance from the bound
protein compared to ChIP-Seq. This step is unique to ChIP-exo and
could potentially provide significantly higher spatial resolution
compared to ChIP-Seq. Finally, high throughput sequencing of a small
region (36 to 100 bp) at the $5\prime$ end of each fragment generates
millions of reads. Figure \ref{fig:chip_diagram} illustrates the
differences between ChIP-exo, Single End (SE) ChIP-Seq and Paired End
(PE) ChIP-Seq: The $5\prime$ ends of a ChIP-exo experiment are located
more tightly around the binding proteins than in a ChIP-Seq
experiment; in a PE ChIP-Seq experiment both ends are observed while
in a SE ChIP-Seq experiment only the $5\prime$ end.

While the number of ChIP-exo data keeps increasing, characteristics of
ChIP-exo data are not fully investigated yet. First, DNA libraries
generated by the ChIP-exo protocol seem to be less complex than the
libraries generated by ChIP-Seq (Mahony et al., 2015
\cite{exo_review}). Second, although there are roughly the same amount
of reads in both strands, locally there may be more reads in one
strand than in the other. Finally, most of current ChIP-Seq quality
control (QC) guidelines (Landt et al., \cite{encode_qc}) may not be
applicable on ChIP-exo, while there are not established QC pipelines
for ChIP-exo; previous ChIP-exo analyses used ChIP-Seq samples to
compare the resolution between experiments (\cite{exo1},
\cite{exoillumina}, \cite{exo2})). To address these challenges, we
suggest a collection of quality control visualizations to interrogate
these biases in a ChIP-exo experiment and globally asses the
enrichment and library complexity of a ChIP-exo sample. We gathered
ChIP-exo data from diverse organisms: CTCF factor in human
\cite{exo1}; ER factor in human and FoxA1 factor in mouse (Serandour
et al., 2013 \cite{exoillumina}); Glucocorticoid receptor (GR) in
IMR90, K562 and U2OS cell lines (Starick et al., 2015
\cite{starick15}); TATA-box Binding Protein (TBP) in K562 cell lines
(Venters and Pugh, 2013 \cite{venters13}). Furthermore, we also
generated ChIP-exo and ChIP-Seq data for $\sig$ factor in
\emph{Escherichia Coli} (\emph{E. Coli}) measured under aerobic ($ +
O_2$) condition, and treated by rifampicin by 0 and 20 minutes
(courtesy of Professor Robert Landick's lab). Additionally we gathered
the ChIP-nexus from TATA-box Binding Protein (TBP) in K562 cell lines;
MyC and Max in S2 cell lines in \emph{Drosophila Melanogaster}
(\emph{D. Melanogaster}); Twist and Dorsal in \emph{D. Melanogaster}
embryo cell lines (He et al., 2015 \cite{chipnexus}).

In order to obtain the potential benefits of ChIP-exo on protein
binding site identification, it is critical to use algorithms that
could fully utilize information available in ChIP-exo data. Rhee and
Pugh, 2011 \cite{exo1} discussed that reads in the forward and reverse
strand might construct peak pairs around bound proteins, of which
heights were implicitly assumed to be symmetric. Based on this
rationale, they used the ``peak pair method'' that predicts the
midpoint of two modes of peak pairs as potential binding
sites. Recently developed ChIP-exo data analysis methods, such as Mace
(Wang et al, 20114 \cite{mace}), CexoR (Madrigal, 2015 \cite{cexor})
and Peakzilla (Bardet et al., 2013 \cite{peakzilla}), are also based
on this peak pair assumption. However, appropriateness of such
assumption was not fully evaluated in the literature yet. Furthermore,
it is still unknown which factors could affect protein binding site
identification using ChIP-exo data. In order to address this problem,
we investigated various aspects of ChIP-exo data by contrasting them
with their respective ChIP-Seq experiments.

Currently, research on statistical methods for ChIP-exo data is still
in its very early stage. Although many methods have been proposed to
identify protein binding sites from ChIP-Seq data (reviewed by
Wilbanks and Facciotti, 2012 \cite{evaluation} and Pepke and Wold,
2009 \cite{computation}), such as MACS (Zhang et al., 2008
\cite{macs}), CisGenome (Ji et al., 2008 \cite{cisgenome}) and MOSAiCS
(Kuan et al., 2009 \cite{mosaics}), these approaches might not fully
utilize potentials of ChIP-exo data for high resolution identification
of protein binding sites. Specifically these approaches reveal protein
binding sites only in lower resolution, i.e., at an interval of
hundreds to thousands of base pairs. Furthermore, they implicitly
assume that there is only one ``mode'' or ``predicted binding
location'' per this wide genomic interval. More recently,
deconvolution algorithms such as Deconvolution (Lun et al., 2009
\cite{csdeconv}), GEM (Guo et al., 2012 \cite{gem}, an improved
version of Guo et al., 2010 \cite{gps} ) and PICS (Zhang et al., 2010
\cite{pics}) have been proposed to identify binding sites in higher
resolution using ChIP-Seq data. However, most of them are still not
tailored for ChIP-exo and PE and SE ChIP-Seq data in a unified
framework and as a result, currently available methods are not
appropriate for fair comparison between ChIP-exo and ChIP-Seq. To
address these limitations, we developed and utilized an improved
version of dPeak (Chung et al., 2013 \cite{dpeak}), a high resolution
binding site identification (deconvolution) algorithm that we
previously developed for PE and SE ChIP-Seq data, so that it can also
handle ChIP-exo data. The dPeak algorithm implements a probabilistic
model that accurately describes the ChIP-exo and ChIP-Seq data
generation process.

Some of the key findings in this work are as follows. First, we
demonstrate that the ``peak pair'' assumption of Rhee and Pugh, 2013
\cite{exo2} does not hold well in real ChIP-exo data. Second, we found
that when we analyze ChIP-exo data from eukaryotic genomes, it is
important to consider sequence biases inherent to ChIP-exo data, such
as mappability and GC content, in order to improve sensitivity and
specificity of binding site identification. Third, we evaluated
several methods to identify binding events and dPeak performs
competitively respect to GEM and MACE when analyzing ChIP-exo
data. Finally, when comparable number of reads is used for both
ChIP-exo and ChIP-Seq , dPeak coupled with ChIP-exo data provides
resolution comparable to PE ChIP-Seq and both significantly improve
the resolution of protein binding identification compared to SE-based
analysis with any of the available methods.

\section{Results and discussion}
\label{sec:results}

\subsection{Deeply sequenced E. Coli $\sig$ ChIP-exo and ChIP-Seq data}

$\sig$ factor is a transcription initiation factor of housekeeping
genes in E. Coli. In this organism's genomes, many promoters contain
multiple transcription start sites (TSS) and these TSS are often
closely spaced (10 $\sim$ 150 bp). These closely spaced binding sites
are considered to be multiple ``switches'' that differentially
regulate gene expression under diverse growth conditions
\cite{regulondb}. Therefore, investigation of ChIP-exo's potential for
identification and differentiation of closely spaced binding sites is
invaluable for elucidating the transcriptional networks of prokaryotic
genomes.

\subsection{Quality Control of ChIP-exo and ChIP-nexus data}

\subsubsection{Application of current ChIP-Seq QC guidelines on ChIP-exo
  data}

We started our exploration by investigating whether the current
state-of-the-art QC pipelines for ChIP-Seq are suitable for
ChIP-exo. Table \ref{tab:qc} contains measures that are commonly
calculated for ChIP-Seq samples: PCR Bottleneck Coefficient (PBC) and
the Normalized Strand Cross-Correlation (NSC) are calculated as in
\cite{encode_qc}. We omitted the Relative Strand Cross-Correlation
(RSC) which is another commonly used QC measure because a typical
ChIP-exo experiment is not accompanied by an input sample.

\begin{table}[h!]
%% need to fix format
  \centering
%% need to re-build this table
<<table1,include = FALSE,echo = FALSE,eval = TRUE>>=
  library(data.table)
  load("../data/paper_tables/Landick_rif_summary.RData")
  load("../data/paper_tables/Landick_aero_summary.RData")
  load("../data/paper_tables/pugh_ctcf.RData")
  load("../data/paper_tables/Carroll_mm9_FoxA1.RData")
  load("../data/paper_tables/Meijsing_hg19_GR.RData")
  load("../data/paper_tables/Carroll_hg19_ER.RData")
  load("../data/paper_tables/Venters_hg19_TBP.RData")
  load("../data/paper_tables/ChIPnexus_dm3.RData")
  load("../data/paper_tables/ChIPnexus_hg19.RData")
  landick_aero[,Organism := "E.Coli"]
  landick_aero[,Replicate := c(1,2,1,2)]
  landick_aero[,TF := "Sig70"]
  landick_aero[ ,cond := c(rep("Aerobic",2),rep("Anaerobic", 2))]
  landick_rif[,Organism := "E.Coli"]
  landick_rif[,Replicate := c(1,1,2,2)]
  landick_rif[,TF := "Sig70"]
  landick_rif[ ,cond := rep(c("Rif-0min","Rif-20min"),each = 2)]
  pugh_ctcf[,Organism := "Human"]
  pugh_ctcf[,Replicate := 1]
  pugh_ctcf[,TF := "CTCF"]
  pugh_ctcf[ ,cond := "HeLA"]
  carroll_mm9_FoxA1[,Organism := "Mouse"]
  carroll_mm9_FoxA1[,Replicate := c(3,1,2)]
  carroll_mm9_FoxA1[,TF := "Fox A1"]
  carroll_mm9_FoxA1[ ,cond := "Mouse Liver"]
  meijing_hg19_GR[,Organism := "Human"]
  meijing_hg19_GR[,Replicate := 1]
  meijing_hg19_GR[, TF := "GR"]
  meijing_hg19_GR[,cond := c("IMR90","K562","U20S")]
  carroll_hg19_ER[,Organism := "Human"]  
  carroll_hg19_ER[,Replicate := c(1,3,2)]
  carroll_hg19_ER[,TF := "ER"]
  carroll_hg19_ER[ ,cond := "MCF-7"]
  venters_hg19_TBP[,Organism := "Human"]
  venters_hg19_TBP[,cond := "K562"]
  venters_hg19_TBP[,Replicate := 1:3]
  venters_hg19_TBP[,TF := "TBP"]
  chipnexus_dm3[, Organism := "D. Melanogaster"]
  chipnexus_dm3[, cond := c(rep("Embryo",4),rep("S2",4))]
  chipnexus_dm3[, Replicate := rep(1:2,4)]
  chipnexus_dm3[, TF := c(rep("Dorsal",2),rep("Twist",2),rep("Max",2),rep("MyC",2))]
  chipnexus_K562[ , Organism := "Human"]
  chipnexus_K562[ , cond  := "K562"]
  chipnexus_K562[ , Replicate := 1:2]
  chipnexus_K562[ , TF := "TBP"]
  ## this creates the table
  A <- do.call(rbind,list(landick_aero,
                     landick_rif[order(Replicate)],
                     pugh_ctcf[order(Replicate)],
                     carroll_mm9_FoxA1[order(Replicate)],
                     meijing_hg19_GR[order(Replicate)],
                     carroll_hg19_ER[order(Replicate)],
                     venters_hg19_TBP[order(Replicate)]))
  B <- do.call(rbind,list(chipnexus_dm3,
                     chipnexus_K562))
  format_tab <- function(A){
    A[,nsc := unlist(nsc)]
    setcolorder(A, c("Organism","TF","cond","Replicate","nreads","pbc","nsc","files"))
    A[,files := NULL]
    A[,nreads := prettyNum(nreads,big.mark = ",")]
    A[,pbc := round(pbc,4)]
    A[,nsc := round(nsc,4)]
    setnames(A,names(A),c("Organism","IP/TF","Condition/Cell","Rep.","Depth","PBC","NSC"))
    return(A)}

<<table2,include = TRUE,echo = FALSE, eval = TRUE >>= 
  A
@ 
\caption{\textbf{Current QC metrics applied to ChIP-exo data.} PBC
  stands for PCR Bottleneck Coefficient and NSC for Normalized Strand
  Cross-Correlation.}
  \label{tab:qc} %% need to make this table
\end{table}

DNA libraries generated by the ChIP-exo protocol seem to be less
complex than the libraries generated by ChIP-Seq, since the possible
number of positions to which the reads can be aligned is being reduced
due to the exonuclease digestion, hence considerable amounts of reads
are being mapped to specific positions. This affects the
interpretation of the PBC, since for ChIP-Seq low PBC values indicate
that the same read has been copied by the amplification process and
aligned multiple times to the same position; while for ChIP-exo when
several reads are aligned to the same position are not necessarily the
same read amplified, but several reads that their $5\prime$ end was
digested to the same position before the amplification step. It is of
special importance to notice that for several ChIP-exo datasets, the
PBC values are quite low. Therefore, by blindly following ChIP-Seq
guidelines those experiment could have been considered as not useful
and repeated.

\subsubsection{Comparison with ChIP-Seq data}

@@

>>>>

\usepackage{xcolor}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{float}
\usepackage{verbatim}
\usepackage{tabu}
\usepackage{multirow}

@@

\section*{Background}
\label{sec:intro}

ChIP-exo (Chromatin Immunoprecipitation followed by exonuclease
digestion and next generation sequencing) Rhee and Pugh, 2011
\nocite{exo1} is the state-of-the-art experiment developed to attain
single base-pair resolution of protein binding site identification and
it is considered as a powerful alternative to popularly used ChIP-Seq
(Chromatin Immunoprecipitation coupled with next generation
sequencing) assay. ChIP-exo experiments first capture millions of DNA
fragments (150 - 250 bp in length) that the protein under study
interacts with using random fragmentation of DNA and a
protein-specific antibody. Then, exonuclease is introduced to trim the
$5\prime$ end of each DNA fragment to a fixed distance from the bound
protein compared to ChIP-Seq. This step is unique to ChIP-exo and
could potentially provide significantly higher spatial resolution
compared to ChIP-Seq. Finally, high throughput sequencing of a small
region (36 to 100 bp) at the $5\prime$ end of each fragment generates
millions of reads. Figure \ref{fig:chip_diagram} illustrates the
differences between ChIP-exo, Single End (SE) ChIP-Seq and Paired End
(PE) ChIP-Seq: The $5\prime$ ends of a ChIP-exo experiment are located
more tightly around the binding proteins than in a ChIP-Seq
experiment; in a PE ChIP-Seq experiment both ends are observed while
in a SE ChIP-Seq experiment only the $5\prime$ end.

ChIP-nexus (Chromatin Immunoprecipitation followed by exonuclease
digestion, unique barcode, single ligation and next generation
ligation) He et al., 2015 \nocite{chipnexus} is a modification to the
ChIP-exo protocol, where both sequencing adaptors are ligated at the
end of the ChIP fragments. Then, after exonuclease digestion, DNA
self-circularization with circLigase, and restriction enzyme cutting
between the two adaptors, the final library is ammplified. ChIP-nexus
arises as an alternative to ChIP-exo as it provides the possibility of
attaining higher resolution analysis and yields higher complexity
libraries.

While the number of ChIP-exo data keeps increasing, characteristics of
ChIP-exo data are not fully investigated yet. First, DNA libraries
generated by the ChIP-exo protocol are expected to be less complex than the
libraries generated by ChIP-Seq (Mahony et al.,
2015\nocite{exo_review}). Second, although there are roughly the same
amount of reads in both strands, locally there may be more reads in
one strand than in the other. Finally, most of current ChIP-Seq
quality control (QC) guidelines (Landt et al., 2012\nocite{encode_qc})
may not be applicable on ChIP-exo, while there are not established QC
pipelines for ChIP-exo; previous ChIP-exo analyses used ChIP-Seq
samples to compare the resolution between experiments
\cite{exo1,exoillumina,exo2}. To address these challenges, we suggest
a collection of quality control visualizations to interrogate these
biases in a ChIP-exo experiment and globally asses the enrichment and
library complexity of a ChIP-exo sample and a procedure to distinguish
low complexity libraries from deeply sequenced experiments. This
aspect is unique to ChIP-exo, since the exonuclease enzyme it is
expected to digested the reads that are not bound to a transcription
factor, therefore the number of bases where a ChIP-exo fragment could
be potentially aligned is reduced. That way, in a high quality
ChIP-exo experiment is possible to observe a large amount of reads to
be aligned to unique position due to genomic signal instead of PCR
artifacts.

In order to obtain the potential benefits of ChIP-exo on protein
binding site identification, it is critical to use algorithms that
could fully utilize information available in ChIP-exo data. Rhee and
Pugh, 2011 \nocite{exo1} discussed that reads in the forward and reverse
strand might construct peak pairs around bound proteins, of which
heights were implicitly assumed to be symmetric. Based on this
rationale, they used the ``peak pair method'' that predicts the
midpoint of two modes of peak pairs as potential binding
sites. Recently developed ChIP-exo data analysis methods, such as Mace
(Wang et al, 20114\nocite{mace}), CexoR (Madrigal, 2015\nocite{cexor})
and Peakzilla (Bardet et al., 2013\nocite{peakzilla}), are also based
on this peak pair assumption. However, appropriateness of such
assumption was not fully evaluated in the literature yet. Furthermore,
it is still unknown which factors could affect protein binding site
identification using ChIP-exo data. In order to address this problem,
we investigated various aspects of ChIP-exo data by contrasting them
with their respective ChIP-Seq experiments.

Currently, research on statistical methods for ChIP-exo data is still
in its very early stage. Although many methods have been proposed to
identify protein binding sites from ChIP-Seq data (reviewed by
Wilbanks and Facciotti, 2012\nocite{evaluation} and Pepke and Wold,
2009 \nocite{computation}), such as MACS (Zhang et al.,
2008\nocite{macs}), CisGenome (Ji et al., 2008\nocite{cisgenome}) and
MOSAiCS (Kuan et al., 2009\nocite{mosaics}), these approaches might
not fully utilize potentials of ChIP-exo data for high resolution
identification of protein binding sites. Specifically these approaches
reveal protein binding sites only in lower resolution, i.e., at an
interval of hundreds to thousands of base pairs. Furthermore, they
implicitly assume that there is only one ``mode'' or ``predicted
binding location'' per this wide genomic interval. More recently,
deconvolution algorithms such as Deconvolution (Lun et al.,
2009\cite{csdeconv}), GEM (Guo et al., 2012\nocite{gem}, an improved
version of Guo et al., 2010\nocite{gps} ) and PICS (Zhang et al.,
2010\nocite{pics}) have been proposed to identify binding sites in
higher resolution using ChIP-Seq data. However, most of them are still
not tailored for ChIP-exo and PE and SE ChIP-Seq data in a unified
framework and as a result, currently available methods are not
appropriate for fair comparison between ChIP-exo and ChIP-Seq. To
address these limitations, we developed and utilized an improved
version of dPeak (Chung et al., 2013\nocite{dpeak}), a high resolution
binding site identification (deconvolution) algorithm that we
previously developed for PE and SE ChIP-Seq data, so that it can also
handle ChIP-exo data. The dPeak algorithm implements a probabilistic
model that accurately describes the ChIP-exo and ChIP-Seq data
generation process.

Some of the key findings in this work are as follows. First, we
demonstrate that the ``peak pair'' assumption of Rhee and Pugh,
2013\nocite{exo2} does not hold well in real ChIP-exo data. Second, we
found that when analyzing ChIP-exo data and the Input control is not
available, it is useful to adjust for GC content and mappability
biases to improve peak calling and binding site identification. Third,
we evaluated several methods to identify binding events and dPeak
performs competitively respect to GEM and MACE when analyzing ChIP-exo
data. Finally, when comparable number of reads is used for both
ChIP-exo and ChIP-Seq , dPeak coupled with ChIP-exo data provides
resolution comparable to PE ChIP-Seq and both significantly improve
the resolution of protein binding identification compared to SE-based
analysis with any of the available methods.

\section*{Results and discussion}
\label{sec:results}
% CTCF factor in human \cite{exo1}; ER factor in
% human and FoxA1 factor in mouse (Serandour et al., 2013
% \cite{exoillumina}); Glucocorticoid receptor (GR) in IMR90, K562 and
% U2OS cell lines (Starick et al., 2015 \cite{starick15}); TATA-box
% Binding Protein (TBP) in K562 cell lines (Venters and Pugh, 2013
% \cite{venters13}). Furthermore, we also generated ChIP-exo and
% ChIP-Seq data for $\sig$ factor in \emph{Escherichia Coli}
% (\emph{E. Coli}) measured under aerobic ($ + O_2$) condition, and
% treated by rifampicin by 0 and 20 minutes (courtesy of Professor
% Robert Landick's lab). Additionally we gathered the ChIP-nexus from
% TATA-box Binding Protein (TBP) in K562 cell lines; MyC and Max in S2
% cell lines in \emph{Drosophila Melanogaster} (\emph{D. Melanogaster});
% Twist and Dorsal in \emph{D. Melanogaster} embryo cell lines (He et
% al., 2015 \cite{chipnexus}).
\subsection*{Publicly available and novel datasets}

We generated $\sig$ factor ChIP-exo, PE and SE ChIP-Seq experiments in
\emph{E. Coli} under aerobic ($+O_2$) and anaerobic ($-O_2$)
conditions in glucose minimal media on the HiSeq2000 and Illumina GA
IIx platforms. Similarly, we generated $\sig$ factor ChIP-exo, PE and
SE (generated \emph{in silico}) ChIP-Seq experiments in \emph{E. Coli}
under aerobic ($+O_2$) conditions where rifampicin was applied for 20
minutes and a control sample (without rifampicin being applied). We
used these experimental designs for comparisons of ChIP-exo and PE
ChIP-Seq assays of high resolution analysis and binding site
identification. This comparison benefits from using $\sig$ for this
study since is a transcription initiation factor of housekeeping genes
in \emph{E. Coli}. In this organism's genomes, many promoters contain
multiple transcription start sites (TSS) and these TSS are often
closely spaced (10 $\sim$ 150 bp). These closely spaced binding sites
are considered to be multiple ``switches'' that differentially
regulate gene expression under diverse growth conditions
\cite{regulondb}.

Additionally, we gathered ChIP-exo data from diverse organisms: CTCF
factor in human HeLa cell lines \cite{exo1}; ER factor in human MCF-7
cell lines \cite{exoillumina}; GR factor in IMR90, K562 and U2OS human
cell lines \cite{starick15}; TBP factor in human K562 cell lines
\cite{venters13}. Additionally, we also gathered the ChIP-nexus
experiments provided by He et al., 2014: TBP in human K562 cell lines,
MyC and Max in S2 \emph{D. Melanogaster} cell lines and, Twist and
Dorsal in \emph{D. Melanogaster} embryo cell lines.

% Therefore, investigation of ChIP-exo's potential for
% identification and differentiation of closely spaced binding sites is
% invaluable for elucidating the transcriptional networks of prokaryotic
% genomes.

@@

\begin{enumerate}
\item Available data, the motivation of this part is two fold:
%  \begin{itemize}
%  \item we want to list all data sets we obtained
%  \item we want to emphasize that for the sig70 datasets we have a
%    unique design that allows us to compare chip exo with SE and PE
%    chipseq , the later one for the first time
%  \end{itemize}
\item comparison with chip seq %? with and without dpeak ?
%\item only in methods. list the align algorithm used
\item chip seq qc metrics
\item chip exo qc pipeline, description and justification of the steps
\begin{itemize}
\item need to justify the sampling step. for this we may use the depth
  table and add the factor to normalize one depth to another. explain
  the confounding of pcr amplification and the reduction of the number
  of possible mappable positions.
\item case study with FoxA1, analysis of enrichment vs library
  complexity and analysis and the strand imbalance tools. in methods
  we need to add which tests did we perform to show that the pipeline
  works. how do we expect from the diagnostics and how to interpret
  the plots.
\item analysis of why the pipeline works. in methods need to specify
  how and in results need to specify why it works, and what are the
  conclusions from it
\end{itemize}
\item high resolution analysis with chip exo and comparison with PE
  chip seq
  \begin{itemize}
  \item justify what do we used dpeak (in methods need to list the
    other methods used). perhaps we can add gof plots in
    supplement. we also want to compare with the same method,
    therefore we can use the fact that dpeak have a unified model for
    PE and SE dpeak
  \item after comparing with other methods we can compare with PE and
    SE ChIP-seq, this comparison gives also resolution and sensitivity
  \item compare samples under same budget of reads
  \end{itemize} 
\end{enumerate}

@@

\subsection*{Comparison with ChIP-Seq data}

@@

\subsection*{Application of ENCODE quality metrics to ChIP-exo and
  ChIP-nexus data}

We started our exploration by investigating whether the current
state-of-the-art QC pipelines for ChIP-Seq are suitable for ChIP-exo
and ChIP-neuxs. In Tables \ref{tab:qc_sig} and \ref{tab:qc} we
calculated a collection of the commonly used ChIP-Seq QC metrics using
the ChIP-exo and ChIP-nexus experiments instead: Normalized Strand
Cross-Correlation (NSC), Relative Strand Cross-Correlation and PCR
Bottleneck Coefficient (PBC) defined as in 
\url{https://genome.ucsc.edu/ENCODE/qualityMetrics.html}\nocite{encode_qc}.

\arrayrulewidth=.1mm

\begin{table}[h!]
  \centering
  \begin{tabu} to\linewidth{X[-1]|X[-1]|X[-1]|X[-1]|X[-1]|X[-1]|X[-1]}
    \firsthline
    \textbf{Condition} & \textbf{Treatment} & \textbf{Rep.} & \textbf{Depth} & \textbf{NSC} & \textbf{RSC} & 
    \textbf{PBC}\\
    \hline 
    Aerobic & No Rif. & 1 & 13,961,493 & 103.15 & 2.0193 & 0.1399 \\
    Aerobic & No Rif. & 2 & 14,810,838 & 162.70 & 1.7805 & 0.1633 \\
    Anaerobic & No Rif. & 1 &  16,108,774 & 153.51 & 1.8035 & 0.1353 \\
    Anaerobic & No Rif. & 2 &  13,636,541 & 172.59 & 2.014 & 0.1532 \\ 
    Aerobic & No Rif. & 1 & 902,921 & 13.77 & 1.1270 & 0.2689\\
    Aerobic & No Rif. & 1 & 1,852,124 & 17.91 & 1.5275 & 0.2590\\
    Aerobic & Rif. 20 min & 2 & 2,104,427 & 29.60 & 1.2844 & 0.2584\\
    Aerobic & Rif. 20 min & 2 & 11,548,572 & 13.08 & 1.5122 & 0.1510 \\
    \lasthline
    \end{tabu}
    \caption{Current QC metrics applied to generated \emph{E. Coli} $\sig$ samples. NSC stands for Normalized 
             Strand Cross-Correlation, RSC stands for Relative Strand Cross-Correlation and PBC stands for 
             PCR Bottleneck Coefficient.}
\label{tab:qc_sig}
\end{table}

\begin{table}[h!]
  \centering
  \begin{tabu} to\linewidth{X[1.3]|X[2.2]|X[-1]|X[-1]|X[-1]|X[1.8]|X[-1]|X[-1]|X[-1]}
    \firsthline
    \textbf{Protocol} & \textbf{Organism} & \textbf{TF} & \textbf{Cell line} & \textbf{Rep.} &
    \textbf{Depth} & \textbf{NSC} & \textbf{RSC} &   \textbf{PBC} \\ 
    \hline
    \multirow{13}{4em}{ChIP-exo} & Human & CTCF & HeLa & & 48,478,450 & 16.02 & 1.1960 & 0.4579 \\
    & \multirow{3}{4em}{Human} & \multirow{3}{4em}{ER} & \multirow{3}{4em}{MCF-7} & 1 & 9,289,835 & 
    19.87 & 1.0127 & 0.8082 \\
    &  &  & & 2 & 11,041,833 & 21.48& 1.0063 & 0.8024\\
    &  &  & & 3 & 12,464,836 & 18.72& 1.0100 & 0.8203 \\    
    &  \multirow{3}{4em}{Mouse} & \multirow{3}{4em}{FoxA1} & \multirow{3}{4em}{Liver} & 1 & 22,210,461 & 21.28 & 1.1104 &  0.6562 \\
    &  & & & 2 & 23,307,557 & 60.42 & 1.1604 & 0.7996 \\ 
    &  & & & 3 & 22,421,72  & 72.04 & 1.1975 & 0.1068 \\
    & \multirow{3}{4em}{Human} & \multirow{3}{4em}{GR} & IM90 & & 47,443,803 & 8.86  & 1.3678 & 0.2978 \\
    & &  & K562 & & 116,518,000   & 4.11 & 1.0441 &0.0504 \\
    & &  & U2OS & & 3,255,111 &  10.05 & 1.0288 & 0.7714 \\
    & \multirow{3}{4em}{Human} & \multirow{3}{4em}{TBP} & \multirow{3}{4em}{K562} &1 & 61,046,382 &  12.01  & 1.1119
 & 0.1232 \\
    & & & &2 & 94,314,770 & 7.93 & 1.0299 & 0.1681 \\
    & & & & 3 & 114,282,270 & 9.25 & 1.1027 & 0.1464\\
\hline
\multirow{10}{4em}{ChIP-nexus} & \multirow{8}{4em}{D.Melanogaster} &
       \multirow{2}{4em}{Dorsal} & \multirow{4}{4em}{embryo} & 1 &   8,863,170 &  7.27 & 1.0402 &  0.6766 \\
 & & &  & 2 & 10,003,562 & 7.19 & 1.0672 & 0.5656\\
 & &  \multirow{2}{4em}{Twist} & & 1 & 18,244,203 & 5.82 &  1.1637 &  0.6592 \\
 & & &  & 2 & 52,546,982 & 5.27  &  1.1805  & 0.4549 \\
 & & \multirow{2}{4em}{Max} & \multirow{4}{4em}{S2} & 1 & 18,320,743 & 3.60 & 1.3628 & 0.5178  \\
 & & & & 2 & 24,965,642  & 3.47 & 1.0138 & 0.2124\\
 & & \multirow{2}{4em}{MyC} & & 1 & 7,832,034 & 5.92 & 1.0115 &  0.3935  \\
 & & & &  2 & 22,824,467 & 5.76 & 1.0045 & 0.1879\\
 & \multirow{2}{4em}{Human} & \multirow{2}{4em}{TBP} & \multirow{2}{4em}{K562} & 1 &  
              33,708,245 & 32.16 & 1.1712 & 0.3102 \\
 & &  &  & 2 & 129,675,001 & 32.70 &  1.2455 & 0.0492 \\
    \lasthline
  \end{tabu}
  \caption{Current QC metrics applied to gathered data. NSC stands for Normalized 
    Strand Cross-Correlation, RSC stands for Relative Strand Cross-Correlation and PBC stands for 
    PCR Bottleneck Coefficient.}  
\label{tab:qc}
\end{table}
